---
title: "SpaceMarkers Functions Step-by-Step"
author: "Atul Deshpande"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: yes
        toc_float:
          toc_collapsed: true
vignette: >
    %\VignetteIndexEntry{SpaceMarkers Functions Step-by-Step}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global.options, include = FALSE}
knitr::opts_knit$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = 'center'
)
knitr::opts_chunk$set(out.extra = 'style="display:block; margin:auto;"')
```

# Overview

This vignette mirrors the *SpaceMarkers with SpatialExperiment Objects* vignette
but replaces the single `SpaceMarkers()` call with each constituent function
called explicitly. Every step dispatches on the `SpaceMarkersExperiment` (SME)
and returns an updated SME, so the analysis is a chain of SME-in / SME-out
calls — no manual slot assignment or pre-extracted matrices required.

The same breast cancer Visium dataset is used throughout.

# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SpaceMarkers")
```

```{r message=FALSE, warning=FALSE}
library(SpaceMarkers)
```

# Setup

## Links to Data

```{r}
data_dir <- "visiumBrCA"
unlink(file.path(data_dir), recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE)
main_10xlink <- "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0"
counts_folder <- "Visium_Human_Breast_Cancer"
counts_file <- "Visium_Human_Breast_Cancer_filtered_feature_bc_matrix.h5"
counts_url <- paste(c(main_10xlink, counts_folder, counts_file), collapse = "/")
sp_folder <- "Visium_Human_Breast_Cancer"
sp_file <- "Visium_Human_Breast_Cancer_spatial.tar.gz"
sp_url <- paste(c(main_10xlink, sp_folder, sp_file), collapse = "/")
```

## Downloading Data

```{r}
download.file(counts_url, file.path(data_dir, basename(counts_url)), mode = "wb")
download.file(sp_url, file.path(data_dir, basename(sp_url)), mode = "wb")
untar(file.path(data_dir, basename(sp_url)), exdir = file.path(data_dir))
```

# Loading Data into a SpaceMarkersExperiment

`load10X()` searches for expression and spatial files with flexible regex
matching, log-transforms counts, and returns an SME with pre-computed kernel
parameters in `spatial_params()`.

```{r message=FALSE, warning=FALSE}
cogaps_result <- readRDS(system.file("extdata", "CoGAPS_result.rds",
    package = "SpaceMarkers", mustWork = TRUE))

sme <- load10X(data_dir, features = cogaps_result, method = "CoGAPS")
sme
```

## Filtering

Retain a curated gene list and spots expressed in at least 3 cells:

```{r}
data("curated_genes")
good_gene_threshold <- 3
keep_genes <- rownames(sme)[
    apply(assay(sme, "logcounts"), 1, function(x) sum(x > 0) >= good_gene_threshold)
]
keep_genes <- intersect(keep_genes, curated_genes)
sme <- sme[keep_genes, ]
sme
```

# Accessors

The SME holds everything the pipeline needs. A few accessors are
worth knowing before walking through the steps:

| Accessor | Returns |
|----------|---------|
| `assay(sme, "logcounts")` | Genes × spots expression matrix |
| `spatialCoords(sme)` | Spot coordinates |
| `spatial_patterns(sme)` | Pattern values per spot |
| `spatial_params(sme)` | Pre-computed kernel parameters (2 × n) |

The SME pipeline methods read these internally — you do not need to extract
them by hand before each step.

```{r}
spatial_params(sme)
```

# Undirected Analysis

The undirected analysis identifies genes that are differentially expressed in
the overlap zone between two patterns. We use Pattern\_1 (immune) and
Pattern\_2, which show substantial spatial co-occurrence in this dataset.

Each step of the pipeline is an SME-in / SME-out method: the SME is passed
through a chain of step functions that each read their inputs from the object
and write their outputs back to the appropriate slot. Writes happen via the
setter generics (`hotspots<-`, `interactions<-`, `undirected_scores<-`,
`overlap_scores<-`, `analysis_type<-`); no manual `@spacemarkers` or
`metadata()` assignment is needed.

```{r}
sme_ud <- sme
spatial_patterns(sme_ud) <- spatial_patterns(sme_ud)[, c("Pattern_1", "Pattern_2")]
spatial_params(sme_ud)   <- spatial_params(sme_ud)[, c("Pattern_1", "Pattern_2"),
                                                   drop = FALSE]
```

## The chain

`find_all_hotspots()` labels each spot relative to each pattern's kernel
smoothing null distribution. `get_pairwise_interacting_genes()` tests each
gene for differential expression between the overlap zone of every pattern
pair and the two exclusive zones. `get_im_scores()` aggregates the per-gene
test results into a single score matrix. `calculate_overlap_undirected()`
quantifies how much each pair's hotspot regions spatially overlap, and sets
`analysis_type(sme) <- "undirected"`.

```{r message=FALSE, warning=FALSE}
sme_ud <- sme_ud |>
    find_all_hotspots() |>
    get_pairwise_interacting_genes(mode = "DE", analysis = "enrichment",
                                   minOverlap = 10, workers = 1) |>
    get_im_scores() |>
    calculate_overlap_undirected()
```

## Inspect the result

```{r}
# Hotspot labels per spot / per pattern
head(hotspots(sme_ud, "undirected"))
```

```{r}
# Fraction of spots classified as hotspots per pattern
hs <- hotspots(sme_ud, "undirected")
pat_cols <- setdiff(colnames(hs), c("barcode", "y", "x"))
vapply(pat_cols, function(p) mean(!is.na(hs[[p]])), numeric(1))
```

```{r}
# Detailed per-gene statistics for the first pattern pair
pair1 <- interactions(sme_ud)[[1]]
if (length(pair1$interacting_genes) > 0) {
    head(pair1$interacting_genes[[1]])
}
```

```{r}
head(undirected_scores(sme_ud))
overlap_scores(sme_ud)
analysis_type(sme_ud)
```

# Directed Analysis

The directed analysis quantifies how much each pattern's *spatial influence*
drives gene expression in the hotspot of the opposing pattern. We use
Pattern\_1 (immune) and Pattern\_5 (cancer), following the main vignette.

```{r}
sme_dir <- sme
spatial_patterns(sme_dir) <- spatial_patterns(sme_dir)[, c("Pattern_1", "Pattern_5")]
spatial_params(sme_dir)   <- spatial_params(sme_dir)[, c("Pattern_1", "Pattern_5"),
                                                     drop = FALSE]
```

## The chain

The directed workflow uses Gaussian mixture model thresholds rather than the
kernel null distribution used in the undirected workflow. `calculate_influence()`
applies kernel smoothing to each pattern independently, producing an
*influence* map — for each spot, how strongly does the pattern radiate into
the neighborhood. `find_hotspots_gmm()` dispatches on an SME and accepts a
`type` argument to choose between operating on the pattern columns
(`type = "pattern"`) or on the computed influence map (`type = "influence"`).
`calculate_gene_scores_directed()` scores each gene per source→target pair
using the intersection of the target's hotspot and the source's influence
zone. `calculate_overlap_directed()` finishes by quantifying hotspot vs.
influence-zone overlap and setting `analysis_type(sme) <- "directed"`.

```{r message=FALSE, warning=FALSE}
sme_dir <- sme_dir |>
    calculate_influence() |>
    find_hotspots_gmm(type = "pattern") |>
    find_hotspots_gmm(type = "influence") |>
    calculate_gene_scores_directed() |>
    calculate_overlap_directed()
```

## Inspect the result

```{r}
head(hotspots(sme_dir, "pattern"))
```

```{r}
head(hotspots(sme_dir, "influence"))
```

```{r}
head(influence_map(sme_dir))
```

```{r}
head(directed_scores(sme_dir))
overlap_scores(sme_dir)
analysis_type(sme_dir)
```

# Visualization

Since results are stored in the SME, all visualization functions work
identically to the high-level workflow.

## IM Scores Bar Plot

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_im_scores(sme_ud, interaction = "Pattern_1_Pattern_2", nGenes = 10)
```

## Spatial Pattern Plots

`plot_spatial()` resolves the feature name from `colData`, gene expression, or
hotspot metadata automatically, and loads the tissue image at the requested
resolution.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_1", title = "Pattern_1 (immune)")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_5", title = "Pattern_5 (cancer)")
```

## Top Interacting Genes

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
ims <- undirected_scores(sme_ud)
top_genes <- head(ims[order(ims$Pattern_1_Pattern_2, decreasing = TRUE), "Gene"], 5)

for (g in top_genes) {
    print(plot_spatial(sme_ud, feature_col = g, title = g))
}
```

# Cleanup

```{r}
unlink(file.path(data_dir), recursive = TRUE)
```

# Session Info

```{r}
sessionInfo()
```
