---
title: "SpaceMarkers Functions Step-by-Step"
author: "Atul Deshpande"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: yes
        toc_float:
          toc_collapsed: true
vignette: >
    %\VignetteIndexEntry{SpaceMarkers Functions Step-by-Step}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global.options, include = FALSE}
knitr::opts_knit$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = 'center'
)
knitr::opts_chunk$set(out.extra = 'style="display:block; margin:auto;"')
```

# Overview

This vignette mirrors the *SpaceMarkers with SpatialExperiment Objects* vignette
but replaces the single `SpaceMarkers()` call with each constituent function
called explicitly. The goal is to show what the pipeline is doing at every step
and how the `SpaceMarkersExperiment` (SME) object serves as both the data source
(via accessors) and the result container (via slot assignment).

The same breast cancer Visium dataset is used throughout.

# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SpaceMarkers")
```

```{r message=FALSE, warning=FALSE}
library(SpaceMarkers)
```

# Setup

## Links to Data

```{r}
data_dir <- "visiumBrCA"
unlink(file.path(data_dir), recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE)
main_10xlink <- "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0"
counts_folder <- "Visium_Human_Breast_Cancer"
counts_file <- "Visium_Human_Breast_Cancer_filtered_feature_bc_matrix.h5"
counts_url <- paste(c(main_10xlink, counts_folder, counts_file), collapse = "/")
sp_folder <- "Visium_Human_Breast_Cancer"
sp_file <- "Visium_Human_Breast_Cancer_spatial.tar.gz"
sp_url <- paste(c(main_10xlink, sp_folder, sp_file), collapse = "/")
```

## Downloading Data

```{r}
download.file(counts_url, file.path(data_dir, basename(counts_url)), mode = "wb")
download.file(sp_url, file.path(data_dir, basename(sp_url)), mode = "wb")
untar(file.path(data_dir, basename(sp_url)), exdir = file.path(data_dir))
```

# Loading Data into a SpaceMarkersExperiment

`load10X()` searches for expression and spatial files with flexible regex
matching, log-transforms counts, and returns an SME with pre-computed kernel
parameters in `spatial_params()`.

```{r message=FALSE, warning=FALSE}
cogaps_result <- readRDS(system.file("extdata", "CoGAPS_result.rds",
    package = "SpaceMarkers", mustWork = TRUE))

sme <- load10X(data_dir, features = cogaps_result, method = "CoGAPS")
sme
```

## Filtering

Retain a curated gene list and spots expressed in at least 3 cells:

```{r}
data("curated_genes")
good_gene_threshold <- 3
keep_genes <- rownames(sme)[
    apply(assay(sme, "logcounts"), 1, function(x) sum(x > 0) >= good_gene_threshold)
]
keep_genes <- intersect(keep_genes, curated_genes)
sme <- sme[keep_genes, ]
sme
```

# Extracting Inputs from the SME

The individual SpaceMarkers functions expect three objects that are each
directly derived from the SME:

| Object | Role | SME accessor |
|--------|------|--------------|
| `expr` | Genes × spots expression matrix | `assay(sme, "logcounts")` |
| `spPatterns` | barcode + coords + pattern columns | `spatialCoords()` + `spatial_patterns()` |
| `optParams` | 2 × n kernel parameter matrix | `spatial_params(sme)` |

```{r}
# Expression matrix (genes x spots)
expr <- assay(sme, "logcounts")
dim(expr)
```

```{r}
# spPatterns: barcode, y, x, then one column per pattern
coords <- as.data.frame(spatialCoords(sme))   # columns: y, x
spPatterns_full <- data.frame(
    barcode = colnames(sme),
    coords,
    as.data.frame(spatial_patterns(sme))
)
rownames(spPatterns_full) <- spPatterns_full$barcode
head(spPatterns_full[, 1:5])
```

```{r}
# Pre-computed kernel parameters (stored by load10X)
# Rows: sigmaOpt, threshOpt  |  Columns: one per pattern
optParams <- spatial_params(sme)
optParams
```

# Undirected Analysis

The undirected analysis identifies genes that are differentially expressed in
the overlap zone between two patterns. We use Pattern\_1 (immune) and
Pattern\_2, which show substantial spatial co-occurrence in this dataset.

```{r}
sme_ud <- sme
spatial_patterns(sme_ud) <- spatial_patterns(sme_ud)[, c("Pattern_1", "Pattern_2")]

# Refresh spPatterns and optParams to match the two-pattern subset
spPatterns_ud <- data.frame(
    barcode = colnames(sme_ud),
    as.data.frame(spatialCoords(sme_ud)),
    as.data.frame(spatial_patterns(sme_ud))
)
rownames(spPatterns_ud) <- spPatterns_ud$barcode

optParams_ud <- spatial_params(sme_ud)
```

## Step 1: Find Per-Pattern Hotspots

`find_all_hotspots()` loops over every pattern column in `spPatterns`, applies
a kernel smoothing null distribution, and labels each spot as a hotspot or NA.
The result is a data.frame with the same barcode/coord columns plus one logical
column per pattern.

```{r message=FALSE, warning=FALSE}
hs <- find_all_hotspots(spPatterns_ud, params = optParams_ud)
head(hs)
```

The fraction of spots classified as hotspots per pattern:

```{r}
pat_cols <- setdiff(colnames(hs), c("barcode", "y", "x"))
vapply(pat_cols, function(p) mean(!is.na(hs[[p]])), numeric(1))
```

## Step 2: Find Pairwise Interacting Genes

`get_pairwise_interacting_genes()` tests each gene for differential expression
between the overlap zone (spots where both patterns are hot) and each pattern's
exclusive zone. It returns a named list — one element per pattern pair — each
containing the per-gene test statistics.

```{r message=FALSE, warning=FALSE}
spaceMarkersResult <- get_pairwise_interacting_genes(
    data      = expr,
    optParams = optParams_ud,
    spPatterns = spPatterns_ud,
    hotspots   = hs,
    mode       = "DE",
    analysis   = "enrichment",
    minOverlap = 10,
    workers    = 1
)
# One list entry per pattern pair
names(spaceMarkersResult)
```

```{r}
# Detailed per-gene statistics for the first pair
pair1 <- spaceMarkersResult[[1]]
if (length(pair1$interacting_genes) > 0) {
    head(pair1$interacting_genes[[1]])
} else {
    message("No interacting genes found for this pair.")
}
```

## Step 3: Calculate Interaction (IM) Scores

`get_im_scores()` aggregates the per-gene test results into a single score
matrix: one row per gene, one column per pattern pair.

```{r}
IMScores <- get_im_scores(spaceMarkersResult)
head(IMScores)
```

## Step 4: Calculate Overlap Scores

`calculate_overlap_undirected()` quantifies how much the hotspot regions of
the two patterns spatially overlap.

```{r}
overlap_ud <- calculate_overlap_undirected(hs)
overlap_ud
```

## Storing Results in the SME

Assign each result to the appropriate slot so that the standard SME accessors
work downstream.

```{r}
# Results go in the spacemarkers$results sub-list
sme_ud@spacemarkers$results <- list(
    undirected_scores = IMScores,
    overlap_scores    = overlap_ud
)
sme_ud@spacemarkers$analysis <- "undirected"

# Detailed intermediates go in metadata()
S4Vectors::metadata(sme_ud)$hotspots$undirected <- hs
S4Vectors::metadata(sme_ud)$interactions        <- spaceMarkersResult

# Verify accessors now work
analysis_type(sme_ud)
head(undirected_scores(sme_ud))
```

# Directed Analysis

The directed analysis quantifies how much each pattern's *spatial influence*
drives gene expression in the hotspot of the opposing pattern. We use
Pattern\_1 (immune) and Pattern\_5 (cancer), following the main vignette.

```{r}
sme_dir <- sme
spatial_patterns(sme_dir) <- spatial_patterns(sme_dir)[, c("Pattern_1", "Pattern_5")]

spPatterns_dir <- data.frame(
    barcode = colnames(sme_dir),
    as.data.frame(spatialCoords(sme_dir)),
    as.data.frame(spatial_patterns(sme_dir))
)
rownames(spPatterns_dir) <- spPatterns_dir$barcode

optParams_dir <- spatial_params(sme_dir)
```

## Step 1: Find Pattern Hotspots (GMM)

The directed workflow uses Gaussian mixture model thresholds rather than the
kernel null distribution used in the undirected workflow. `calculate_thresholds()`
fits a two-component GMM to each pattern's value distribution and returns the
boundary between the low and high components. `find_hotspots_gmm()` then labels
each spot above that boundary as a hotspot.

```{r message=FALSE, warning=FALSE}
patthresholds <- calculate_thresholds(
    spPatterns_dir, minvals = 0.1, maxvals = 0.8
)
patthresholds
```

```{r message=FALSE, warning=FALSE}
patHotspots <- find_hotspots_gmm(spPatterns_dir, threshold = patthresholds)
head(patHotspots)
```

Fraction of spots classified as hotspots per pattern:

```{r}
pat_cols_dir <- setdiff(colnames(patHotspots), c("barcode", "y", "x"))
vapply(pat_cols_dir, function(p) mean(!is.na(patHotspots[[p]])), numeric(1))
```

## Step 2: Calculate Spatial Influence

`calculate_influence()` applies kernel smoothing to each pattern independently,
producing an *influence* map: for each spot, how strongly does the pattern
radiate into the neighborhood? The returned data.frame has the same
barcode/coord columns as `spPatterns` plus one influence column per pattern.

```{r message=FALSE, warning=FALSE}
spInfluence <- calculate_influence(spPatterns_dir, optParams_dir)
head(spInfluence)
```

## Step 3: Find Influence Hotspots (GMM)

The same GMM approach is applied to the influence values to identify spots
that fall under strong influence of each pattern.

```{r message=FALSE, warning=FALSE}
infthresholds <- calculate_thresholds(
    spInfluence, minvals = 0.01, maxvals = 0.5
)
infthresholds
```

```{r message=FALSE, warning=FALSE}
infHotspots <- find_hotspots_gmm(spInfluence, threshold = infthresholds)
head(infHotspots)
```

## Step 4: Calculate Directed Gene Scores

`calculate_gene_scores_directed()` scores each gene for each directed pattern
pair (source → target). Spots in the *target* pattern's hotspot that also fall
under the *source* pattern's influence define the interaction zone.

```{r}
patnames <- setdiff(colnames(spPatterns_dir), c("barcode", "x", "y"))
patternPairs <- t(combn(patnames, 2))
patternPairs
```

```{r message=FALSE, warning=FALSE}
IMscores_dir <- calculate_gene_scores_directed(
    data               = expr,
    pat_hotspots       = patHotspots,
    influence_hotspots = infHotspots,
    pattern_pairs      = patternPairs
)
head(IMscores_dir)
```

## Step 5: Calculate Directed Overlap Scores

`calculate_overlap_directed()` quantifies how much each pattern's hotspot
overlaps with the opposing pattern's *influence* zone.

```{r}
overlap_dir <- calculate_overlap_directed(patHotspots, infHotspots)
overlap_dir
```

## Storing Results in the SME

```{r}
sme_dir@spacemarkers$results <- list(
    directed_scores = IMscores_dir,
    overlap_scores  = overlap_dir
)
sme_dir@spacemarkers$analysis <- "directed"

S4Vectors::metadata(sme_dir)$hotspots$pattern   <- patHotspots
S4Vectors::metadata(sme_dir)$hotspots$influence  <- infHotspots
S4Vectors::metadata(sme_dir)$influence            <- spInfluence

# Verify accessors now work
analysis_type(sme_dir)
head(directed_scores(sme_dir))
```

```{r}
# Intermediate intermediates are accessible via convenience functions
head(hotspots(sme_dir, type = "pattern"))
head(hotspots(sme_dir, type = "influence"))
head(influence_map(sme_dir))
```

# Visualization

Since results are stored in the SME, all visualization functions work
identically to the high-level workflow.

## IM Scores Bar Plot

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_im_scores(sme_ud, interaction = "Pattern_1_Pattern_2", nGenes = 10)
```

## Spatial Pattern Plots

`plot_spatial()` resolves the feature name from `colData`, gene expression, or
hotspot metadata automatically, and loads the tissue image at the requested
resolution.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_1", title = "Pattern_1 (immune)")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_5", title = "Pattern_5 (cancer)")
```

## Top Interacting Genes

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
ims <- undirected_scores(sme_ud)
top_genes <- head(ims[order(ims$Pattern_1_Pattern_2, decreasing = TRUE), "Gene"], 5)

for (g in top_genes) {
    print(plot_spatial(sme_ud, feature_col = g, title = g))
}
```

# Cleanup

```{r}
unlink(file.path(data_dir), recursive = TRUE)
```

# Session Info

```{r}
sessionInfo()
```
