---
title: "SpaceMarkers RCTD"
author: "Orian Stapleton"
date: "2024-01-08"
output:
    BiocStyle::html_document:
        toc: yes
        toc_depth: 3
        number_sections: yes
vignette: >
    %\VignetteIndexEntry{Inferring Immune Interactions in Breast Cancer}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
params:
  workdir: "."
---

```{r global.options, include = FALSE}
knitr::opts_knit$set(
    collapse = TRUE,
    comment = "#>",
    fig.align   = 'center'
)
knitr::opts_knit$set(
  root.dir = params$workdir
  )

knitr::opts_chunk$set(out.extra = 'style="display:block; margin:auto;"')

```

# Overview

SpaceMarkers leverages latent feature analysis of the spatial components of 
transcriptomic data to identify biologically relevant molecular interactions 
between cell groups.This tutorial will use the latent features from CoGAPS to 
look at pattern interactions in a Visium 10x breast ductal carcinoma spatial 
transcriptomics dataset.

# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SpaceMarkers")
```

```{r message = FALSE, warning = FALSE}
library(SpaceMarkers)
library(spacexr)
```


# RCTD

## Spatial Setup

```{r}
main_10xlink <- "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0"
counts_folder <- "Visium_Human_Breast_Cancer"
counts_file <- "Visium_Human_Breast_Cancer_filtered_feature_bc_matrix.h5"
counts_url<-paste(c(main_10xlink,counts_folder,counts_file), collapse = "/")
sp_folder <- "Visium_Human_Breast_Cancer"
sp_file <- "Visium_Human_Breast_Cancer_spatial.tar.gz"
sp_url <- paste(c(main_10xlink,sp_folder,sp_file),collapse = "/")
## Clear previous runs
unlink(basename(sp_url))
unlink("spatial", recursive = TRUE)
files <- list.files(".")[grepl(basename(counts_url),list.files("."))]
unlink(files)

## Counts Matrix
download.file(counts_url,basename(counts_url), mode = "wb")
counts_matrix <- load10XExpr(visiumDir = ".", h5filename = counts_file)
good_gene_threshold <- 3
goodGenes <- rownames(counts_matrix)[
    apply(counts_matrix,1,function(x) sum(x>0)>=good_gene_threshold)]
spCounts <- counts_matrix[goodGenes,]

## Coords
download.file(sp_url, basename(sp_url), mode = "wb")
untar(basename(sp_url))
spCoords <- load10XCoords(visiumDir = ".")
rownames(spCoords) <- spCoords$barcode
spCoords <- spCoords[,-1]

## Clear Folders
unlink(basename(sp_url))
unlink("spatial", recursive = TRUE)
files <- list.files(".")[grepl(basename(counts_url),list.files("."))]
unlink(files)
```


## scRNA-seq Reference Setup

```{r eval = FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(spacexr)
#devtools::install_github("dmcable/spacexr", build_vignettes = FALSE)

# Load the  dataset
scObj <- Read10X(data.dir = "RCTD/scRNASeqRef_BreastCancer/")
meta <- read.csv("RCTD/scRNASeqRef_BreastCancer/Whole_miniatlas_meta.csv")
meta <- meta[-1,]
rownames(meta) <- meta[,1]
scObj <- CreateSeuratObject(counts = scObj, project = "BreastCancer",meta.data = meta)
# saveRDS(scObj,'RCTD/scRef_BreastCancer.rds')
# saveRDS(meta,'RCTD/scRef_meta.rds')
# scObj <- readRDS("RCTD/scRef_BreastCancer.rds")
# meta <- readRDS('RCTD/scRef_meta.rds') 

#scRNA Reference
celltypes <- meta$celltype_major
names(celltypes) <- rownames(meta)
celltypes <- as.factor(celltypes)
features <- intersect(rownames(scObj@assays$RNA$counts),goodGenes)
scCounts <- scObj@assays$RNA$count[features,]
nUMI <- colSums(scCounts)
reference <- Reference(counts = scCounts, cell_types = celltypes,nUMI = nUMI,
                       require_int = FALSE, n_max_cells = 35000)
#saveRDS(reference, "RCTD/RCTD_reference.rds")
#Spatial RNA
puck <- SpatialRNA(spCoords[,c("x","y")], spCounts,nUMI = NULL,
                   require_int = FALSE)
#saveRDS(puck, "RCTD/RCTD_spatial.rds")

```

## Run RCTD

```{r eval = FALSE}
# RCTD - ran on the cluster - jhpce01 (takes about 5 minutes with these settings)
library(spacexr) 
#reference <- readRDS('data/visiumBreastCancer_reference.rds')
#puck <- readRDS('data/visiumBreastCancer_RCTDspatial.rds')
myRCTD <- create.RCTD(puck, reference, max_cores = 8)
myRCTD <- run.RCTD(myRCTD, doublet_mode = 'full')
saveRDS(myRCTD,'outputs/RCTD_visiumBreastCancer.rds')
```


# SpaceMarkers

## Spatial Features

```{r}
spFeatures <- getSpatialFeatures("RCTD/RCTD_visiumBreastCancer.rds",method = "RCTD")
spFeatures <- spacexr::normalize_weights(spFeatures)
```

## Spatial Patterns

```{r}
spCoords$barcode <- rownames(spCoords)
barcodes <- intersect(rownames(spCoords),rownames(spFeatures))
spPatterns <- cbind(spCoords[barcodes,],as.data.frame(as.matrix(
    spFeatures[barcodes,])))
head(spPatterns)
```

## optParams

```{r eval = FALSE}
optParams <- getSpatialParameters(spPatterns)

```
```{r}
optParams <- readRDS("RCTD/optParams.rds") 
```

