---
title: "SpaceMarkers with SpatialExperiment Objects"
author: "Atul Deshpande"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: yes
        toc_float:
          toc_collapsed: true
vignette: >
    %\VignetteIndexEntry{SpaceMarkers with SpatialExperiment Objects}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global.options, include = FALSE}
knitr::opts_knit$set(
    collapse = TRUE,
    comment = "#>",
    fig.align   = 'center'
)

knitr::opts_chunk$set(out.extra = 'style="display:block; margin:auto;"')

```

# Overview

This vignette demonstrates how to use the `SpaceMarkersExperiment` class, a
`SpatialExperiment` subclass that serves as a unified container for all
SpaceMarkers inputs and outputs. Instead of tracking separate matrices, data
frames, and lists across pipeline steps, a single `SpaceMarkersExperiment`
object holds everything: expression data, spatial coordinates, pattern
features, optimal parameters, hotspots, interacting genes, and interaction
scores.

We will reproduce the same breast cancer analysis from the main vignette using
this integrated workflow.

# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SpaceMarkers")
```
```{r message = FALSE, warning = FALSE}
library(SpaceMarkers)
```

# Setup

## Links to Data

```{r}
data_dir <- "visiumBrCA"
unlink(file.path(data_dir), recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE)
main_10xlink <- "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0"
counts_folder <- "Visium_Human_Breast_Cancer"
counts_file <- "Visium_Human_Breast_Cancer_filtered_feature_bc_matrix.h5"
counts_url <- paste(c(main_10xlink, counts_folder, counts_file), collapse = "/")
sp_folder <- "Visium_Human_Breast_Cancer"
sp_file <- "Visium_Human_Breast_Cancer_spatial.tar.gz"
sp_url <- paste(c(main_10xlink, sp_folder, sp_file), collapse = "/")
```

## Downloading Data

```{r}
download.file(counts_url, file.path(data_dir, basename(counts_url)), mode = "wb")
download.file(sp_url, file.path(data_dir, basename(sp_url)), mode = "wb")
untar(file.path(data_dir, basename(sp_url)), exdir = file.path(data_dir))
```

# Loading Data into a SpaceMarkersExperiment

## Using `load10X`

`load10X()` searches the Visium directory for expression and spatial files
using flexible pattern matching — it works regardless of whether your files are
in an `outs/`, `data/`, or flat directory layout. It log-transforms the counts
and returns a `SpaceMarkersExperiment` directly:

```{r message=FALSE, warning=FALSE}
cogaps_result <- readRDS(system.file("extdata", "CoGAPS_result.rds",
    package = "SpaceMarkers", mustWork = TRUE))

sme <- load10X(data_dir, features = cogaps_result, method = "CoGAPS")
sme
```

Alternatively, you can load first and add features separately with
`add_features()`:

```{r eval=FALSE}
# Step 1: Load without features
sme <- load10X(data_dir)

# Step 2: Add features later
sme <- add_features(sme, cogaps_result, method = "CoGAPS")
```

If the features cover fewer spots than the expression data (or vice versa),
`add_features()` will keep only the shared spots and emit a warning.

The object inherits from `SpatialExperiment` so all standard accessors work:

```{r}
# Expression matrix
dim(assay(sme, "logcounts"))

# Spatial coordinates
head(spatialCoords(sme))

# Spatial patterns stored in colData
head(spatial_patterns(sme))
```

## Filtering

For this demonstration we will use two patterns and a curated gene list:

```{r}
data("curated_genes")
good_gene_threshold <- 3
keep_genes <- rownames(sme)[
    apply(assay(sme, "logcounts"), 1, function(x) sum(x > 0) >= good_gene_threshold)
]
keep_genes <- intersect(keep_genes, curated_genes)
sme <- sme[keep_genes, ]

# Keep all patterns for now; we will subset per analysis below
sme
```

# Running SpaceMarkers with the SME Object

## Undirected Analysis

The `SpaceMarkers()` function now accepts a `SpaceMarkersExperiment` as
its first argument. All results are stored back in the returned object.
For the undirected analysis we use Pattern_1 and Pattern_2, which show
substantial overlap in this dataset.

```{r message=FALSE, warning=FALSE}
sme_ud <- sme
spatial_patterns(sme_ud) <- spatial_patterns(sme_ud)[, c("Pattern_1", "Pattern_2")]

sme_undirected <- SpaceMarkers(
    x = sme_ud,
    directed = FALSE,
    min.gene.expr = 3
)
```

### Inspecting Results

Summary scores are accessed via dedicated accessors:

```{r}
# Analysis type
analysis_type(sme_undirected)

# IM scores (summary: genes x pattern pairs)
head(undirected_scores(sme_undirected))
```

Detailed intermediate results are stored in `metadata()` and accessed via
convenience functions:

```{r}
# Undirected hotspots
hs <- hotspots(sme_undirected, type = "undirected")
head(hs)
```

```{r}
# Per-pair interaction details
pair_names <- names(interactions(sme_undirected))
pair_names

# Detailed stats for the first pair (if any interactions were found)
if (length(pair_names) > 0) {
    pair1 <- interactions(sme_undirected, pair = pair_names[1])
    if (length(pair1$interacting_genes) > 0) {
        head(pair1$interacting_genes[[1]])
    } else {
        message("No interacting genes found for this pair.")
    }
}
```

## Directed Analysis

For the directed analysis we use Pattern_1 (immune) and Pattern_5 (cancer),
following the main SpaceMarkers vignette.

```{r message=FALSE, warning=FALSE}
sme_dir <- sme
spatial_patterns(sme_dir) <- spatial_patterns(sme_dir)[, c("Pattern_1", "Pattern_5")]

sme_directed <- SpaceMarkers(
    x = sme_dir,
    directed = TRUE,
    min.gene.expr = 3
)
```

```{r}
analysis_type(sme_directed)
head(directed_scores(sme_directed))
```

Directed analysis also stores hotspot and influence intermediates:

```{r}
# Pattern hotspots
head(hotspots(sme_directed, type = "pattern"))

# Influence hotspots
head(hotspots(sme_directed, type = "influence"))

# Influence map
head(influence_map(sme_directed))
```

# Visualization

Since the SME holds all results, every visualization function accepts the
object directly and reads from the stored slots.

## Patterns on the tissue

`plot_spatial()` automatically extracts coordinates from the object, finds
the tissue image stored in the object's `visiumDir`, and falls back to a
blank background if none is available. The `source` argument selects where
`feature_col` is resolved: `colData` (patterns, continuous features),
`assay` (gene expression), `hotspots` (hotspot labels), `influence_map`
(per-pattern influence values), or `interaction` (a three-way categorical
overlay).

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_1", title = "Pattern_1 (immune)")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_5", title = "Pattern_5 (cancer)")
```

## Undirected hotspots and interaction overlay

Pass `source = "hotspots"` so `plot_spatial()` reads hotspot labels from
`metadata(sme)$hotspots$undirected` rather than the continuous pattern
column in `colData` (which shares the same name). `colors` maps the "hot"
level to a single fill.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_undirected, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "undirected",
             colors = "steelblue", title = "Pattern_1 hotspots")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_undirected, feature_col = "Pattern_2",
             source = "hotspots", hotspot_type = "undirected",
             colors = "firebrick", title = "Pattern_2 hotspots")
```

A three-way overlay classifies each hotspot spot as Pattern\_1 only,
interacting (hot in both), or Pattern\_2 only:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_undirected,
             source = "interaction", hotspot_type = "undirected",
             interaction_patterns = c("Pattern_1", "Pattern_2"),
             colors = c(Pattern_1   = "steelblue",
                        interacting = "purple",
                        Pattern_2   = "firebrick"))
```

## IM scores bar plot and top genes

The column names of `undirected_scores(sme)` are the pattern-pair keys —
use the first pair key robustly instead of hard-coding:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
pair_col <- setdiff(colnames(undirected_scores(sme_undirected)), "Gene")[1]
plot_im_scores(sme_undirected, interaction = pair_col, nGenes = 10)
```

`plot_spatial()` accepts gene names directly — it falls through to the
expression assay via `source = "auto"` (the default):

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
ims <- undirected_scores(sme_undirected)
top_genes <- head(ims[order(ims[[pair_col]], decreasing = TRUE), "Gene"], 5)

for (g in top_genes) {
    print(plot_spatial(sme_undirected, feature_col = g, title = g))
}
```

## Undirected overlap scores

`plot_overlap_scores()` reads `overlap_scores(sme)` directly and
auto-detects the undirected shape (symmetric `pattern1 × pattern2`
heatmap):

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_overlap_scores(sme_undirected, title = "Undirected overlap")
```

## Directed hotspots and interaction overlay

The directed workflow splits hotspots into two flavors — pattern-value
GMM hotspots (`hotspot_type = "pattern"`) and influence-map GMM hotspots
(`hotspot_type = "influence"`) — and the interaction overlay's middle
label auto-switches to `"<source> near <target>"`.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "pattern",
             colors = "steelblue", title = "Pattern_1 GMM hotspots")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed, feature_col = "Pattern_5",
             source = "hotspots", hotspot_type = "pattern",
             colors = "firebrick", title = "Pattern_5 GMM hotspots")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed,
             source = "interaction", hotspot_type = "pattern",
             interaction_patterns = c("Pattern_1", "Pattern_5"),
             colors = c(Pattern_1                  = "steelblue",
                        "Pattern_1 near Pattern_5" = "purple",
                        Pattern_5                  = "firebrick"))
```

## Directed overlap scores

`plot_overlap_scores()` detects the directed shape (`pattern × influence`
with a relative-abundance fill) automatically:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_overlap_scores(sme_directed, title = "Directed overlap (relative abundance)")
```

# Constructing a SpaceMarkersExperiment Manually

If you already have expression and coordinate data in R, you can construct
a `SpaceMarkersExperiment` directly:

```{r}
mat <- matrix(rnorm(100), nrow = 10, ncol = 10)
rownames(mat) <- paste0("gene_", 1:10)
colnames(mat) <- paste0("spot_", 1:10)

coords <- matrix(runif(20) * 100, ncol = 2)
colnames(coords) <- c("y", "x")
rownames(coords) <- colnames(mat)

cd <- S4Vectors::DataFrame(
    CellType_A = runif(10),
    CellType_B = runif(10),
    row.names = colnames(mat)
)

my_sme <- SpaceMarkersExperiment(
    assays = list(logcounts = mat),
    colData = cd,
    spatialCoords = coords,
    spaceMarkers = list(
        params = list(pattern_names = c("CellType_A", "CellType_B"))
    )
)
my_sme
```

## From an Existing SpatialExperiment

You can also extract features from any `SpatialExperiment` object using
`get_spatial_features()`:

```{r}
spe <- SpatialExperiment::SpatialExperiment(
    assays = list(logcounts = mat),
    colData = cd,
    spatialCoords = coords
)

# Extract numeric colData columns as spatial features
features <- get_spatial_features(spe)
head(features)
```

# Backward Compatibility

The legacy file-path interface still works. Set `returnSME = FALSE` to get
the old-style output:

```{r eval = FALSE}
# Legacy usage (returns data.frame, not SME)
result <- SpaceMarkers(
    features = "path/to/features.rds",
    data = "path/to/visium_dir",
    directed = FALSE,
    returnSME = FALSE
)
```

# Summary

| Feature | Legacy Workflow | SME Workflow |
|---|---|---|
| Input | File paths | `SpaceMarkersExperiment` or file paths |
| Output | Scattered data.frames/lists | Single `SpaceMarkersExperiment` |
| Expression | Separate matrix | `assay(sme, "logcounts")` |
| Coordinates | Separate data.frame | `spatialCoords(sme)` |
| Patterns | Merged in spPatterns | `spatial_patterns(sme)` |
| Kernel Parameters | Separate matrix | `spatial_params(sme)` |
| All Hyperparameters | Scattered arguments | `params(sme)` |
| Undirected Scores | Separate data.frame | `undirected_scores(sme)` |
| Directed Scores | Separate data.frame | `directed_scores(sme)` |
| LR Scores | Separate matrix | `lr_scores(sme)` |
| Overlap Scores | Separate data.frame | `overlap_scores(sme)` |
| Hotspots | Separate data.frame | `hotspots(sme, type = ...)` |
| Interactions | Nested list | `interactions(sme, pair = ...)` |
| Influence | Separate data.frame | `influence_map(sme)` |
| Spatial Plotting | Manual coord extraction + `plot_spatial_data_over_image` | `plot_spatial(sme, feature_col = ...)` |

# Cleanup

```{r}
unlink(file.path(data_dir), recursive = TRUE)
```

# Session Info

```{r}
sessionInfo()
```
