---
title: "SpaceMarkers Functions Step-by-Step"
author: "Atul Deshpande"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: yes
        toc_float:
          toc_collapsed: true
vignette: >
    %\VignetteIndexEntry{SpaceMarkers Functions Step-by-Step}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global.options, include = FALSE}
knitr::opts_knit$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = 'center'
)
knitr::opts_chunk$set(
    out.extra = 'style="display:block; margin:auto;"',
    warning = FALSE,
    message = FALSE
)
```

# Overview

This vignette mirrors the *SpaceMarkers with SpatialExperiment Objects* vignette
but replaces the single `SpaceMarkers()` call with each constituent function
called explicitly. Every step dispatches on the `SpaceMarkersExperiment` (SME)
and returns an updated SME, so the analysis is a chain of SME-in / SME-out
calls — no manual slot assignment or pre-extracted matrices required.

The same breast cancer Visium dataset is used throughout.

# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("SpaceMarkers")
```

```{r message=FALSE, warning=FALSE}
library(SpaceMarkers)
```

# Setup

## Links to Data

```{r}
data_dir <- "visiumBrCA"
unlink(file.path(data_dir), recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE)
main_10xlink <- "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0"
counts_folder <- "Visium_Human_Breast_Cancer"
counts_file <- "Visium_Human_Breast_Cancer_filtered_feature_bc_matrix.h5"
counts_url <- paste(c(main_10xlink, counts_folder, counts_file), collapse = "/")
sp_folder <- "Visium_Human_Breast_Cancer"
sp_file <- "Visium_Human_Breast_Cancer_spatial.tar.gz"
sp_url <- paste(c(main_10xlink, sp_folder, sp_file), collapse = "/")
```

## Downloading Data

```{r}
download.file(counts_url, file.path(data_dir, basename(counts_url)), mode = "wb")
download.file(sp_url, file.path(data_dir, basename(sp_url)), mode = "wb")
untar(file.path(data_dir, basename(sp_url)), exdir = file.path(data_dir))
```

# Loading Data into a SpaceMarkersExperiment

`load10X()` searches for expression and spatial files with flexible regex
matching, log-transforms counts, and returns an SME with pre-computed kernel
parameters in `spatial_params()`.

```{r message=FALSE, warning=FALSE}
cogaps_result <- readRDS(system.file("extdata", "CoGAPS_result.rds",
    package = "SpaceMarkers", mustWork = TRUE))

sme <- load10X(data_dir, features = cogaps_result, method = "CoGAPS")
sme
```

## Filtering

Retain a curated gene list and spots expressed in at least 3 cells:

```{r}
data("curated_genes")
good_gene_threshold <- 3
keep_genes <- rownames(sme)[
    apply(assay(sme, "logcounts"), 1, function(x) sum(x > 0) >= good_gene_threshold)
]
keep_genes <- intersect(keep_genes, curated_genes)
sme <- sme[keep_genes, ]
sme
```

# Accessors

The SME holds everything the pipeline needs. A few accessors are
worth knowing before walking through the steps:

| Accessor | Returns |
|----------|---------|
| `assay(sme, "logcounts")` | Genes × spots expression matrix |
| `spatialCoords(sme)` | Spot coordinates |
| `spatial_patterns(sme)` | Pattern values per spot |
| `spatial_params(sme)` | Pre-computed kernel parameters (2 × n) |

The SME pipeline methods read these internally — you do not need to extract
them by hand before each step.

```{r}
spatial_params(sme)
```

# Undirected Analysis

The undirected analysis identifies genes that are differentially expressed in
the overlap zone between two patterns. We focus on Pattern\_1 (immune) and
Pattern\_2, which show substantial spatial co-occurrence in this dataset.

Every step of the pipeline is an SME-in / SME-out method: the SME is passed
through a chain of step functions that each read their inputs from the object
and write their outputs back to the appropriate slot. Writes happen via the
setter generics (`hotspots<-`, `interactions<-`, `undirected_scores<-`,
`overlap_scores<-`, `analysis_type<-`); no manual `@spacemarkers` or
`metadata()` assignment is needed.

We'll walk through the five pipeline steps one at a time, visualizing what
each step produces.

## Step 0: Visualize the starting patterns

Before running any pipeline step, inspect the CoGAPS patterns on the tissue.
`plot_spatial()` resolves a feature name against `colData` (for patterns),
`rownames(sme)` (for gene expression), or the hotspot metadata (via the
`hotspot_type` argument).

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_1", title = "Pattern_1 (immune)")
```

And the second pattern we'll use for the undirected analysis:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_2", title = "Pattern_2")
```

## Step 1: Find hotspots

`find_all_hotspots()` labels each spot as a hotspot (or `NA`) for every
pattern, using a Gaussian-kernel null distribution with the pre-computed
`spatial_params(sme)`. The result is stored in
`metadata(sme)$hotspots$undirected` via the `hotspots<-` setter.

```{r message=FALSE, warning=FALSE}
sme_ud <- find_all_hotspots(sme)
```

```{r}
# Fraction of spots classified as hotspots per pattern
hs <- hotspots(sme_ud, "undirected")
pat_cols <- setdiff(colnames(hs), c("barcode", "y", "x"))
vapply(pat_cols, function(p) mean(!is.na(hs[[p]])), numeric(1))
```

Plot the hotspot labels for each pattern — pass `source = "hotspots"` so
`plot_spatial()` reads from the hotspots metadata rather than the continuous
pattern column in `colData`. The `colors` argument controls fill: for
categorical hotspot labels (`"hot"` / `NA`) `plot_spatial()` passes it into
`ggplot2::scale_fill_manual()`, so a single color string maps the `"hot"`
level to that color. Other aesthetic knobs: `point_size`, `stroke`, `alpha`,
`bg_color`, `text_size`, and `crop`.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_ud, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "undirected",
             colors = "steelblue", title = "Pattern_1 hotspots")
```

…and the Pattern\_2 hotspots picked up by the same call (a single
`find_all_hotspots()` populates hotspots for every pattern column):

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_ud, feature_col = "Pattern_2",
             source = "hotspots", hotspot_type = "undirected",
             colors = "firebrick", title = "Pattern_2 hotspots")
```

A combined interaction overlay classifies each hotspot spot as Pattern\_1
only, interacting (hot in both), or Pattern\_2 only. Pass
`source = "interaction"` with `interaction_patterns = c("Pattern_1",
"Pattern_2")` and a three-color vector to distinguish the three classes.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_ud,
             source = "interaction", hotspot_type = "undirected",
             interaction_patterns = c("Pattern_1", "Pattern_2"),
             colors = c(Pattern_1 = "steelblue",
                        interacting = "purple",
                        Pattern_2 = "firebrick"))
```

## Step 2: Find pairwise interacting genes

`get_pairwise_interacting_genes()` tests each gene for differential
expression between the Pattern\_1 × Pattern\_2 overlap zone and each
pattern's exclusive zone. The `pattern_pairs` argument scopes the test to a
single pair so the narrative stays focused; omit it to compute all pairs.
The full per-gene result list is written to `metadata(sme)$interactions`.

```{r message=FALSE, warning=FALSE}
sme_ud <- get_pairwise_interacting_genes(
    sme_ud,
    mode = "DE", analysis = "enrichment",
    minOverlap = 10, workers = 1,
    pattern_pairs = rbind(c("Pattern_1", "Pattern_2"))
)
```

```{r}
# Detailed per-gene statistics for the Pattern_1 x Pattern_2 pair
names(interactions(sme_ud))
pair1 <- interactions(sme_ud)[[1]]
if (length(pair1$interacting_genes) > 0) {
    head(pair1$interacting_genes[[1]])
}
```

## Step 3: Calculate IM scores

`get_im_scores()` aggregates the per-gene Kruskal–Wallis + Dunn's-test
statistics from the previous step into a single score matrix (genes × pattern
pairs), stored via `undirected_scores<-`.

```{r}
sme_ud <- get_im_scores(sme_ud)
head(undirected_scores(sme_ud))
```

The top-ranked genes for the Pattern\_1 × Pattern\_2 interaction:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
pair_col <- setdiff(colnames(undirected_scores(sme_ud)), "Gene")[1]
plot_im_scores(sme_ud, interaction = pair_col, nGenes = 10)
```

Plot the top gene's spatial expression — the IM score is a summary of where
a gene is differentially expressed in the overlap zone.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
ims <- undirected_scores(sme_ud)
top_gene <- ims[order(ims[[pair_col]], decreasing = TRUE), "Gene"][1]
plot_spatial(sme_ud, feature_col = top_gene,
             title = sprintf("%s (top %s gene)", top_gene, pair_col))
```

## Step 4: Calculate overlap scores

`calculate_overlap_undirected()` measures how much each pair's hotspots
actually overlap in space (Szymkiewicz–Simpson by default). The resulting
table is stored via `overlap_scores<-` and `analysis_type(sme)` is set to
`"undirected"`.

```{r}
sme_ud <- calculate_overlap_undirected(sme_ud)
overlap_scores(sme_ud)
analysis_type(sme_ud)
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_overlap_scores(sme_ud)
```

# Directed Analysis

The directed analysis quantifies how much each pattern's *spatial influence*
drives gene expression in the hotspot of the opposing pattern. We focus on
Pattern\_1 (immune) → Pattern\_5 (cancer), matching the main vignette.

The directed workflow uses Gaussian mixture model thresholds rather than the
kernel null distribution used in the undirected workflow, so the hotspot
step differs — you supply `type = "pattern"` (on the pattern values) or
`type = "influence"` (on the smoothed influence map).

## Step 0: Visualize the starting patterns

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_1", title = "Pattern_1 (immune)")
```

The directed analysis focuses on how Pattern\_1 (immune) radiates into
the Pattern\_5 (cancer) region, so we'll track the cancer pattern in the
same way:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_5", title = "Pattern_5 (cancer)")
```

## Step 1: Calculate influence

`calculate_influence()` applies kernel smoothing to each pattern column
independently, producing per-spot values that describe how far each pattern
"radiates" into the surrounding tissue. The result is a data.frame stored in
`metadata(sme)$influence` via `influence_map<-`.

```{r message=FALSE, warning=FALSE}
sme_dir <- calculate_influence(sme)
head(influence_map(sme_dir))
```

Plot the smoothed influence map for each pattern (`source = "influence_map"`
makes `plot_spatial` read from `metadata(sme)$influence` instead of the
continuous pattern column in `colData`):

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_1",
             source = "influence_map",
             title = "Pattern_1 influence")
```

Pattern\_5's influence map, on the same tissue:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_5",
             source = "influence_map",
             title = "Pattern_5 influence")
```

## Step 2: Find pattern hotspots

`find_hotspots_gmm(type = "pattern")` fits a two-component Gaussian mixture
model to each pattern's value distribution and labels spots above the
boundary as hotspots. Stored under `metadata(sme)$hotspots$pattern`.

```{r message=FALSE, warning=FALSE}
sme_dir <- find_hotspots_gmm(sme_dir, type = "pattern")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "pattern",
             colors = "steelblue", title = "Pattern_1 GMM hotspots")
```

…and Pattern\_5's GMM hotspot, which defines the *target* region the
directed analysis will test:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_5",
             source = "hotspots", hotspot_type = "pattern",
             colors = "firebrick", title = "Pattern_5 GMM hotspots")
```

## Step 3: Find influence hotspots

Running `find_hotspots_gmm(type = "influence")` applies the same GMM
thresholding but to the smoothed influence map from Step 1 — spots under
strong influence of each pattern. Stored under
`metadata(sme)$hotspots$influence`.

```{r message=FALSE, warning=FALSE}
sme_dir <- find_hotspots_gmm(sme_dir, type = "influence")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "influence",
             colors = "steelblue", title = "Pattern_1 influence hotspots")
```

Pattern\_5's influence hotspot — the complementary *source* region used
in the reverse-direction test (Pattern\_5 influencing Pattern\_1):

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir, feature_col = "Pattern_5",
             source = "hotspots", hotspot_type = "influence",
             colors = "firebrick", title = "Pattern_5 influence hotspots")
```

With both pattern and influence hotspots in place, the directed
interaction classification is fully defined. It replaces the three-way
undirected scheme (`pat1 / interacting / pat2`) with an asymmetric four-
or five-way factor: `pat1`, `"<pat1> near <pat2>"`, `"<pat2> near <pat1>"`,
`pat2`, and (optionally) `"both"` — the doubly-hot confounder region that
`calculate_gene_scores_directed` drops when `avoid_confounders = TRUE`.
A spot gets `"<P> near <Q>"` when it sits in pattern `P`'s GMM hotspot
AND pattern `Q`'s influence zone, i.e. the exact region the directed
t-test scores.

The classification is exposed via `overlap_map()` for direct use in
custom plots or downstream analysis:

```{r}
labs <- overlap_map(sme_dir, c("Pattern_1", "Pattern_5"))
table(labs, useNA = "ifany")
```

`plot_spatial(..., source = "interaction")` calls `overlap_map()`
internally and auto-picks directed mode when both
`hotspots(sme, "pattern")` and `hotspots(sme, "influence")` are
populated, so the rendered overlay is always in sync with the labels
above:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_dir,
             source = "interaction",
             interaction_patterns = c("Pattern_1", "Pattern_5"),
             colors = c(Pattern_1                  = "steelblue",
                        "Pattern_1 near Pattern_5" = "purple",
                        both                       = "gray40",
                        "Pattern_5 near Pattern_1" = "goldenrod",
                        Pattern_5                  = "firebrick"))
```

## Step 4: Directed gene scores

`calculate_gene_scores_directed()` scores each gene for each source → target
pair, using the intersection of the target's GMM hotspot and the source's
influence hotspot as the interaction zone. `pattern_pairs` restricts to the
Pattern\_1 × Pattern\_5 pair. Results written via `directed_scores<-`.

```{r message=FALSE, warning=FALSE}
sme_dir <- calculate_gene_scores_directed(
    sme_dir,
    pattern_pairs = rbind(c("Pattern_1", "Pattern_5"))
)
```

```{r}
head(directed_scores(sme_dir))
```

The top gene for the strongest directed interaction, plotted on the tissue:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
ds <- directed_scores(sme_dir)
ci <- sort(unique(ds$cell_interaction))[1]
top_gene <- ds[ds$cell_interaction == ci, ]
top_gene <- top_gene[order(-top_gene$effect_size), "gene"][1]
plot_spatial(sme_dir, feature_col = top_gene,
             title = sprintf("%s (top directed gene, %s)", top_gene, ci))
```

## Step 5: Directed overlap scores

`calculate_overlap_directed()` quantifies how much each pattern's hotspot
overlaps with the opposing pattern's influence zone, and sets
`analysis_type(sme)` to `"directed"`.

```{r}
sme_dir <- calculate_overlap_directed(sme_dir)
overlap_scores(sme_dir)
analysis_type(sme_dir)
```

`plot_overlap_scores()` auto-detects whether the stored scores are
undirected (columns `pattern1`, `pattern2`, `overlapScore`) or directed
(columns `pattern`, `influence`, `relAbundance`) and picks the right axes
and fill scale accordingly.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_overlap_scores(sme_dir, title = "Directed overlap (relative abundance)")
```

# Putting it together

Once the individual steps make sense, the whole analysis is a single pipe
chain:

```{r eval=FALSE}
sme_ud <- sme |>
    find_all_hotspots() |>
    get_pairwise_interacting_genes(
        mode = "DE", analysis = "enrichment",
        minOverlap = 10, workers = 1,
        pattern_pairs = rbind(c("Pattern_1", "Pattern_2"))
    ) |>
    get_im_scores() |>
    calculate_overlap_undirected()

sme_dir <- sme |>
    calculate_influence() |>
    find_hotspots_gmm(type = "pattern") |>
    find_hotspots_gmm(type = "influence") |>
    calculate_gene_scores_directed(
        pattern_pairs = rbind(c("Pattern_1", "Pattern_5"))
    ) |>
    calculate_overlap_directed()
```

# Cleanup

```{r}
unlink(file.path(data_dir), recursive = TRUE)
```

# Session Info

```{r}
sessionInfo()
```
