---
title: "SpaceMarkers with SpatialExperiment Objects"
author: "Atul Deshpande"
date: "`r Sys.Date()`"
output:
    BiocStyle::html_document:
        toc: yes
        toc_float:
          toc_collapsed: true
vignette: >
    %\VignetteIndexEntry{SpaceMarkers with SpatialExperiment Objects}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r global.options, include = FALSE}
knitr::opts_knit$set(
    collapse = TRUE,
    comment = "#>",
    fig.align   = 'center'
)

knitr::opts_chunk$set(
    out.extra = 'style="display:block; margin:auto;"',
    warning = FALSE,
    message = FALSE
)

```

# Overview

This vignette demonstrates how to use the `SpaceMarkersExperiment` class, a
`SpatialExperiment` subclass that serves as a unified container for all
SpaceMarkers inputs and outputs. Instead of tracking separate matrices, data
frames, and lists across pipeline steps, a single `SpaceMarkersExperiment`
object holds everything: expression data, spatial coordinates, pattern
features, optimal parameters, hotspots, interacting genes, and interaction
scores.

We will reproduce the same breast cancer analysis from the main vignette using
this integrated workflow.

# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("SpaceMarkers")
```
```{r message = FALSE, warning = FALSE}
library(SpaceMarkers)
library(SpatialExperiment)
```

# Setup

## Links to Data

```{r}
data_dir <- "visiumBrCA"
unlink(file.path(data_dir), recursive = TRUE)
dir.create(data_dir, showWarnings = FALSE)
main_10xlink <- "https://cf.10xgenomics.com/samples/spatial-exp/1.3.0"
counts_folder <- "Visium_Human_Breast_Cancer"
counts_file <- "Visium_Human_Breast_Cancer_filtered_feature_bc_matrix.h5"
counts_url <- paste(c(main_10xlink, counts_folder, counts_file), collapse = "/")
sp_folder <- "Visium_Human_Breast_Cancer"
sp_file <- "Visium_Human_Breast_Cancer_spatial.tar.gz"
sp_url <- paste(c(main_10xlink, sp_folder, sp_file), collapse = "/")
```

## Downloading Data

```{r}
download.file(counts_url, file.path(data_dir, basename(counts_url)), mode = "wb")
download.file(sp_url, file.path(data_dir, basename(sp_url)), mode = "wb")
untar(file.path(data_dir, basename(sp_url)), exdir = file.path(data_dir))
```

# Loading Data into a SpaceMarkersExperiment

## Using `load10X`

`load10X()` searches the Visium directory for expression and spatial files
using flexible pattern matching — it works regardless of whether your files are
in an `outs/`, `data/`, or flat directory layout. It log-transforms the counts
and returns a `SpaceMarkersExperiment` directly:

```{r message=FALSE, warning=FALSE}
cogaps_result <- readRDS(system.file("extdata", "CoGAPS_result.rds",
    package = "SpaceMarkers", mustWork = TRUE))

sme <- load10X(data_dir, features = cogaps_result, method = "CoGAPS")
sme
```

Alternatively, you can load first and add features separately with
`add_features()`:

```{r eval=FALSE}
# Step 1: Load without features
sme <- load10X(data_dir)

# Step 2: Add features later
sme <- add_features(sme, cogaps_result, method = "CoGAPS")
```

If the features cover fewer spots than the expression data (or vice versa),
`add_features()` will keep only the shared spots and emit a warning.

The object inherits from `SpatialExperiment` so all standard accessors work:

```{r}
# Expression matrix
dim(assay(sme, "logcounts"))

# Spatial coordinates
head(spatialCoords(sme))

# Spatial patterns stored in colData
head(spatial_patterns(sme))
```

## Filtering

For this demonstration we will use two patterns and a curated gene list:

```{r}
data("curated_genes")
good_gene_threshold <- 3
keep_genes <- rownames(sme)[
    apply(assay(sme, "logcounts"), 1, function(x) sum(x > 0) >= good_gene_threshold)
]
keep_genes <- intersect(keep_genes, curated_genes)
sme <- sme[keep_genes, ]

# Keep all patterns for now; we will subset per analysis below
sme
```

# Running SpaceMarkers with the SME Object

## Undirected Analysis

The `SpaceMarkers()` function now accepts a `SpaceMarkersExperiment` as
its first argument. All results are stored back in the returned object.
We keep all five CoGAPS patterns on the SME (so the overlap heatmap shows
every pairwise relationship) and restrict the gene-level interaction
tests to Pattern\_1 × Pattern\_2 — our narrative focus — via
`pattern_pairs`, which `SpaceMarkers()` forwards to
`get_pairwise_interacting_genes()`.

```{r message=FALSE, warning=FALSE}
sme_undirected <- SpaceMarkers(
    x = sme,
    directed = FALSE,
    min.gene.expr = 3,
    pattern_pairs = rbind(c("Pattern_1", "Pattern_2"))
)
```

### Inspecting Results

Summary scores are accessed via dedicated accessors:

```{r}
# Analysis type
analysis_type(sme_undirected)

# IM scores (summary: genes x pattern pairs)
head(undirected_scores(sme_undirected))
```

Detailed intermediate results are stored in `metadata()` and accessed via
convenience functions:

```{r}
# Undirected hotspots
hs <- hotspots(sme_undirected, type = "undirected")
head(hs)
```

```{r}
# Per-pair interaction details
pair_names <- names(interactions(sme_undirected))
pair_names

# Detailed stats for the first pair (if any interactions were found)
if (length(pair_names) > 0) {
    pair1 <- interactions(sme_undirected, pair = pair_names[1])
    if (length(pair1$interacting_genes) > 0) {
        head(pair1$interacting_genes[[1]])
    } else {
        message("No interacting genes found for this pair.")
    }
}
```

## Directed Analysis

The directed analysis keeps every pattern on the SME so the directed
overlap heatmap covers all source → target pairs, and restricts
gene-level scoring to Pattern\_1 → Pattern\_5 (immune → cancer) via
`pattern_pairs`, which `SpaceMarkers()` forwards to
`calculate_gene_scores_directed()`.

```{r message=FALSE, warning=FALSE}
sme_directed <- SpaceMarkers(
    x = sme,
    directed = TRUE,
    min.gene.expr = 3,
    pattern_pairs = rbind(c("Pattern_1", "Pattern_5"))
)
```

```{r}
analysis_type(sme_directed)
head(directed_scores(sme_directed))
```

Directed analysis also stores hotspot and influence intermediates:

```{r}
# Pattern hotspots
head(hotspots(sme_directed, type = "pattern"))

# Influence hotspots
head(hotspots(sme_directed, type = "influence"))

# Influence map
head(influence_map(sme_directed))
```

# Visualization

Since the SME holds all results, every visualization function accepts the
object directly and reads from the stored slots.

## Patterns on the tissue

`plot_spatial()` automatically extracts coordinates from the object, finds
the tissue image stored in the object's `visiumDir`, and falls back to a
blank background if none is available. The `source` argument selects where
`feature_col` is resolved: `colData` (patterns, continuous features),
`assay` (gene expression), `hotspots` (hotspot labels), `influence_map`
(per-pattern influence values), or `interaction` (a three-way categorical
overlay).

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_1", title = "Pattern_1 (immune)")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme, feature_col = "Pattern_5", title = "Pattern_5 (cancer)")
```

## Undirected hotspots

Pass `source = "hotspots"` so `plot_spatial()` reads hotspot labels from
`metadata(sme)$hotspots$undirected` rather than the continuous pattern
column in `colData` (which shares the same name). `colors` maps the "hot"
level to a single fill.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_undirected, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "undirected",
             colors = "steelblue", title = "Pattern_1 hotspots")
```

…and Pattern\_2, from the same `find_all_hotspots()` pass:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_undirected, feature_col = "Pattern_2",
             source = "hotspots", hotspot_type = "undirected",
             colors = "firebrick", title = "Pattern_2 hotspots")
```

## Undirected overlap scores

With hotspots in hand we can already answer the broader spatial question
— which pattern pairs actually share territory. `plot_overlap_scores()`
reads `overlap_scores(sme)` directly and auto-detects the undirected
shape (symmetric `pattern1 × pattern2` heatmap) across every pair:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_overlap_scores(sme_undirected, title = "Undirected overlap")
```

## Undirected interaction overlay

For a chosen pair, a three-way overlay classifies each hotspot spot as
Pattern\_1 only, interacting (hot in both), or Pattern\_2 only. This is
the spatial region the per-gene tests below operate on.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_undirected,
             source = "interaction", hotspot_type = "undirected",
             interaction_patterns = c("Pattern_1", "Pattern_2"),
             colors = c(Pattern_1   = "steelblue",
                        interacting = "purple",
                        Pattern_2   = "firebrick"))
```

## IM scores bar plot and top genes

The column names of `undirected_scores(sme)` are the pattern-pair keys —
use the first pair key robustly instead of hard-coding:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
pair_col <- setdiff(colnames(undirected_scores(sme_undirected)), "Gene")[1]
plot_im_scores(sme_undirected, interaction = pair_col, nGenes = 10)
```

`plot_spatial()` accepts gene names directly — it falls through to the
expression assay via `source = "auto"` (the default):

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
ims <- undirected_scores(sme_undirected)
top_genes <- head(ims[order(ims[[pair_col]], decreasing = TRUE), "Gene"], 5)

for (g in top_genes) {
    print(plot_spatial(sme_undirected, feature_col = g, title = g))
}
```

## Directed hotspots

The directed workflow splits hotspots into two flavors: pattern-value
GMM hotspots (`hotspot_type = "pattern"`) and influence-map GMM hotspots
(`hotspot_type = "influence"`). Here we focus on the pattern-level
hotspots that define the source/target regions of each directed pair.

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed, feature_col = "Pattern_1",
             source = "hotspots", hotspot_type = "pattern",
             colors = "steelblue", title = "Pattern_1 GMM hotspots")
```

…and Pattern\_5, the target region the immune pattern radiates into:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed, feature_col = "Pattern_5",
             source = "hotspots", hotspot_type = "pattern",
             colors = "firebrick", title = "Pattern_5 GMM hotspots")
```

## Directed overlap scores

`plot_overlap_scores()` detects the directed shape (`pattern ×
influence` with a relative-abundance fill) automatically, again across
every source → target pair:

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_overlap_scores(sme_directed, title = "Directed overlap (relative abundance)")
```

## Directed interaction overlays

Because a directed analysis runs two independent t-tests per pair — one
per source → target direction — the per-spot classification is best
rendered one direction at a time. The forward view shows Pattern\_1's
hotspot split by whether each spot is also in Pattern\_5's influence
zone (the test's interaction group):

```{r}
labs_fwd <- overlap_map(sme_directed, c("Pattern_1", "Pattern_5"),
                        direction = "forward")
table(labs_fwd, useNA = "ifany")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed,
             source = "interaction",
             interaction_patterns = c("Pattern_1", "Pattern_5"),
             direction = "forward",
             colors = c(Pattern_1                  = "steelblue",
                        "Pattern_1 near Pattern_5" = "purple",
                        "Pattern_5 influence"      = "gray60"))
```

And the reverse (Pattern\_5 → Pattern\_1):

```{r}
labs_rev <- overlap_map(sme_directed, c("Pattern_1", "Pattern_5"),
                        direction = "reverse")
table(labs_rev, useNA = "ifany")
```

```{r, message=FALSE, warning=FALSE, dpi=60, fig.width=6}
plot_spatial(sme_directed,
             source = "interaction",
             interaction_patterns = c("Pattern_1", "Pattern_5"),
             direction = "reverse",
             colors = c(Pattern_5                  = "firebrick",
                        "Pattern_5 near Pattern_1" = "goldenrod",
                        "Pattern_1 influence"      = "gray60"))
```

# Constructing a SpaceMarkersExperiment Manually

If you already have expression and coordinate data in R, you can construct
a `SpaceMarkersExperiment` directly:

```{r}
mat <- matrix(rnorm(100), nrow = 10, ncol = 10)
rownames(mat) <- paste0("gene_", 1:10)
colnames(mat) <- paste0("spot_", 1:10)

coords <- matrix(runif(20) * 100, ncol = 2)
colnames(coords) <- c("y", "x")
rownames(coords) <- colnames(mat)

cd <- S4Vectors::DataFrame(
    CellType_A = runif(10),
    CellType_B = runif(10),
    row.names = colnames(mat)
)

my_sme <- SpaceMarkersExperiment(
    assays = list(logcounts = mat),
    colData = cd,
    spatialCoords = coords,
    spaceMarkers = list(
        params = list(pattern_names = c("CellType_A", "CellType_B"))
    )
)
my_sme
```

## From an Existing SpatialExperiment

You can also extract features from any `SpatialExperiment` object using
`get_spatial_features()`:

```{r}
spe <- SpatialExperiment::SpatialExperiment(
    assays = list(logcounts = mat),
    colData = cd,
    spatialCoords = coords
)

# Extract numeric colData columns as spatial features
features <- get_spatial_features(spe)
head(features)
```

# Backward Compatibility

The legacy file-path interface still works. Set `returnSME = FALSE` to get
the old-style output:

```{r eval = FALSE}
# Legacy usage (returns data.frame, not SME)
result <- SpaceMarkers(
    features = "path/to/features.rds",
    data = "path/to/visium_dir",
    directed = FALSE,
    returnSME = FALSE
)
```

# Summary

| Feature | Legacy Workflow | SME Workflow |
|---|---|---|
| Input | File paths | `SpaceMarkersExperiment` or file paths |
| Output | Scattered data.frames/lists | Single `SpaceMarkersExperiment` |
| Expression | Separate matrix | `assay(sme, "logcounts")` |
| Coordinates | Separate data.frame | `spatialCoords(sme)` |
| Patterns | Merged in spPatterns | `spatial_patterns(sme)` |
| Kernel Parameters | Separate matrix | `spatial_params(sme)` |
| All Hyperparameters | Scattered arguments | `params(sme)` |
| Undirected Scores | Separate data.frame | `undirected_scores(sme)` |
| Directed Scores | Separate data.frame | `directed_scores(sme)` |
| LR Scores | Separate matrix | `lr_scores(sme)` |
| Overlap Scores | Separate data.frame | `overlap_scores(sme)` |
| Hotspots | Separate data.frame | `hotspots(sme, type = ...)` |
| Interactions | Nested list | `interactions(sme, pair = ...)` |
| Influence | Separate data.frame | `influence_map(sme)` |
| Spatial Plotting | Manual coord extraction + `plot_spatial_data_over_image` | `plot_spatial(sme, feature_col = ...)` |

# Cleanup

```{r}
unlink(file.path(data_dir), recursive = TRUE)
```

# Session Info

```{r}
sessionInfo()
```
